Emergent dual scaling of riverine biodiversity

### 2021-03-25 manuscript_pnas.Rmd
Update manuscript_pnas.Rmd

Here, we show a dual scaling law of biodiversity along the two orthogonal dimensions of ecosystem size and complexity in rivers - a prime example of complex branching ecosystems. Individual streams and rivers flow through different landscapes with distinct geological and climatic backgrounds, serving as a spatial unit of unique in-stream environments [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018]. The recurrent merging of diverse tributaries ultimately forms a fluvial network with fractal branching patterns [@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014], may represent a macro-scale control of the ecosystem's habitat heterogeneity (habitat diversity per unit area) [@grantLivingBranchesPopulation2007; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018]. Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity), two factors that regulate biodiversity at the metacommunity level [@hubbellUnifiedNeutralTheory2001; @kallimanisHowDoesHabitat2008]. We predict that ecosystem size and branching complexity increase watershed-scale species richness ($\gamma$ diversity) by enhancing local species richness ($\alpha$ diversity) and/or spatial difference of species composition ($\beta$ diversity) under different ecological scenarios. The present study combines theory and empirical analysis of extensive community data from two different regions of the globe to provide crucial insights into how ecological communities are structured in complex branching networks.

@@

Our theoretical analysis yielded results consistent with our prediction. Ecosystem size and complexity both increased $\gamma$ diversity under a realistic landscape scenario (**Figures 2 and 3**), where the environmental variation at headwaters $\sigma_h$ is greater than the degree of local environmental noise $\sigma_l$ (**Figure 1A**). The power-law relationships (i.e., linear in a log-log scale) have emerged along the both axes of ecosystem size and complexity irrespective of ecological scenarios (compare panels in **Figures 2 and 3**). Importantly, the impact of ecosystem complexity was comparable to that of ecosystem size. Hence, regardless of ecological scenarios, ecosystem size and complexity are likely to be equally important in scaling $\gamma$ diversity.

>>>>

Here, we propose a unified framework of ecosystem size and complexity in scaling biodiversity patterns in rivers - a prime example of complex branching ecosystems. Individual streams and rivers flow through different landscapes with distinct geological and climatic backgrounds, serving as a spatial unit of unique in-stream environments [@brownDendriticNetworkStructure2010; @petersonModellingDendriticEcological2013; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018; @isaakSlowClimateVelocities2016; @mcguireNetworkAnalysisReveals2014; @abbottUnexpectedSpatialStability2018]. The recurrent merging of diverse tributaries ultimately forms a fluvial network with fractal branching patterns [@rodriguez-iturbeFractalRiverBasins2001]. As such, the complexity of branching structure, which we define here as the probability of branching per unit river distance [@teruiMetapopulationStabilityBranching2018; @yeakelSynchronisationStabilityRiver2014], may represent a macro-scale control of the ecosystem's habitat heterogeneity (habitat diversity per unit area) [@grantLivingBranchesPopulation2007; @mooreEmergentStabilityLarge2015; @teruiMetapopulationStabilityBranching2018]. Meanwhile, ecosystem size (watershed area) should determine the metacommunity size and total habitat diversity (area x heterogeneity), two factors that regulate biodiversity at the metacommunity level [@hubbellUnifiedNeutralTheory2001; @kallimanisHowDoesHabitat2008]. The present study combines theory and analysis of extensive community data from two different regions of the globe to show that multiple ecological pathways converge to the emergence of dual diversity scaling along the two orthogonal dimensions of branching networks. Specifically, watershed-scale species richness ($\gamma$ diversity) increased linearly with ecosystem size and branching complexity on logarithmic axes regardless of ecological contexts; however, contributing factors of increased $\gamma$ diversity - i.e., either enhanced local species richness ($\alpha$ diversity) and/or spatial variation of species composition ($\beta$ diversity) - depended on constituent species' characteristics. Our findings suggest that the dual scaling law is a pervasive yet overlooked feature of complex ecosystems with important implications for biodiversity conservation.

@@

Theory predicted that ecosystem size and complexity both increased $\gamma$ diversity under a realistic landscape scenario (**Figures 2 and 3**), where the environmental variation at headwaters $\sigma_h$ is greater than the degree of local environmental noise $\sigma_l$ (**Figure 1A**). The power-law relationships (i.e., linear in a log-log scale) have emerged along the both axes of ecosystem size and complexity irrespective of ecological scenarios (compare panels in **Figures 2 and 3**). Importantly, the impact of ecosystem complexity was comparable to that of ecosystem size. Hence, regardless of ecological scenarios, ecosystem size and complexity are likely to be equally important in scaling $\gamma$ diversity.
